### 2014-04-20 text/Chapter03.tex
Tweaks and additions in bounded reuse (Chapter 3)

\paragraph{Summary.}
The structural coeffect system for tracking liveness analysis

semantics is ``the other way round''

@@

\subsection{Bounded reuse}

Bounded linear logic \cite{girard1992bounded} restricts well-typed terms to polynomial-time algorithms.  
This is done by limiting the number of times a value (proposition) can be used. An assumption
$!_k A$ means that a variable can be used at most $k$ times. 
We attach annotations to the whole context rather than individual assumptions and so a context
$!_{k_1} A_1, ..., !_{k_n} A_n$ is written as $\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$. 
This difference is further explained in Section~\ref{sec:related-work}. %For uniformity
%with later systems, we also write assumptions $A$ as $\tau$ and annotations $K$ as $\cclrd{r}$.

@@

\begin{figure}[t]
{\small Ordinary rules}

\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{1}} \vdash x : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau} 
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \sigma}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, v : \sigma}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \sigma}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau}
\end{equation*}

{\small Structural rules}

\begin{equation*}
\tyrule{weak}
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{0}} \vdash e : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
\end{equation*}
\begin{equation*}
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
\end{equation*}

\caption{Coeffect rules for tracking implicit parameters}
\label{fig:applications-struct-bll}
\vspace{-1em}
\end{figure}

\paragraph{Bounded linear coeffects.}
The system in Figure~\ref{fig:bounded-coeff} extends the outlined idea into a simple calculus.
Variable access (\emph{var}) has a singleton context with a singleton coeffect vector
$\alift{1}$. Weakening (\emph{weak}) extends the free-variable context with an unused variable and 
the coeffect with an associated scalar $0$. Explicit contraction (\emph{contr}) and
exchange (\emph{exch}) rules manipulate variables in the context and
modify the annotations accordingly -- adding the number of uses in
contraction and switching vector elements in exchange.

For abstraction (\emph{abs}), we know the number of uses of the parameter variable $x$ 
and attach it to the function type $\sigma \xrightarrow{\cclrd{s}}
\tau$ as a \emph{latent} coeffect. The remaining variables in $\Gamma$ are 
annotated with the remaining coeffect vector $\aclrd{R}$, specifying \emph{immediate} coeffects.

Application (\emph{app}) describes call-by-name evaluation.  Applying
a function that uses its parameter $\cclrd{t}$-times to an argument
that uses variables in $\Gamma_2$ $\aclrd{S}$-times means that, in
total, the variables in $\Gamma_2$ will be used $(\cclrd{t} \ast \aclrd{S})$-times. 
Recall that $\cclrd{t} \ast \aclrd{S}$ is a scalar multiplication of a
vector. Meanwhile, the variables in $\Gamma_1$ are used just
$\aclrd{R}$-times when reducing the expression $e_1$ to a function
value.

\paragraph{Example.} To demonstrate, consider a term
$(\lambda v.x+v+v)~(x+y)$. According to the call-by-name intuition, the variable $x$ is used three 
times -- once directly inside the function and twice via the variable $v$ after 
substitution. Similarly, $y$ is used 
twice. Assuming a judgment for the function body, abstraction yields:
%
\begin{equation*}
\tyrule{abs}
 { \coctx{x\!:\!\mathbb{Z},v:\mathbb{Z}}{\alift{\cclrd{1},\cclrd{2}}} \vdash x+v+v : \mathbb{Z} }
 { \coctx{x\!:\!\mathbb{Z}}{\alift{\cclrd{1}}} \vdash (\lambda v.x+v+v) : \mathbb{Z} \xrightarrow{\cclrd{2}} \mathbb{Z} }
\end{equation*}
%
To avoid name clashes, we $\alpha$-rename $x$ to $x'$ and later join $x$ and $x'$ using contraction.
Assuming $(x'+y)$ is checked in a context that marks $x'$ and $y$ as used once, the application rule yields
a judgment that is simplified as follows:
\[
\hspace{-0.7em}\inference
  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}
          {\alift{\cclrd{1}} \cons (\cclrd{2} \ast \alift{\cclrd{1},\cclrd{1}}) } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
{\inference
  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{1}, \cclrd{2}, \cclrd{2}} } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
  { \coctx{x\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{3}, \cclrd{2}}} \vdash (\lambda v.x+v+v)~(x+y)  : \mathbb{Z}} }
\]
%
The first step performs scalar multiplication, producing the vector
$\alift{\cclrd{1},\cclrd{2},\cclrd{2}}$. In the second step, we use contraction to join variables 
$x$ and $x'$ from the function and argument terms respectively.

It is worth pointing out that reduction by substitution yields $x+(x+y)+(x+y)$ which has the same coeffect as
the original. We return to evaluation strategies in Section~\ref{sec:syntax}, and 
show that structural coeffect systems preserve types and coeffects under $\beta$-reduction.

>>>>

\paragraph{Summary.}
The structural liveness calculus is a typical example of a system that tracks per-variable 
annotations. In a number of ways, the system is simpler than the flat coeffect calculi. In
lambda abstraction, we simply annotate function with the annotation of a matching variable 
(this rule is the same for all upcoming systems). In application, the \emph{pointwise} composition
is no longer needed, because the sub-expressions use separate contexts. On the other hand, 
we had to add weakening, contraction and exchange rules to let us manipulate the contexts.

The semantics of weakening demonstrates an important point about coeffects that may be quite
confusing. When we read the \emph{typing rule} from top to bottom, weakening adds a variable
to the context. When we read the \emph{semantic rule}, weakening drops a variable value from the 
context! This duality is caused by the fact that coeffects talk about context -- they describe
how to build the context required by the sub-expressions and so the semantics implements 
transformation from the context in the (typing) conclusion to the (typing) assumption.

The structural systems discussed in the upcoming sections are remarkably similar to the one
shown here. We discuss two more examples in details to explore the design space, but we shall
omit details that are the shared with the system in this section.

@@

\subsection{Bounded variable use}

Liveness analysis checks whether a variable is used or unused. With structural coeffects, we can go
further and track how many times is the variable accessed. Girard et al. \cite{logic-bounded} coined
this idea as \emph{bounded linear logic} and use it to restrict well-typed programs to 
polynomial-time algorithms. We first introduce the system in our, coeffect, style and then 
relate it with the original formulation.

\begin{figure}[t]
{\small a.) Ordinary, syntax-driven rules with sub-coeffecting}
\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{\cclrd{1}}} \vdash x : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
\end{equation*}
\vspace{0.5em}

{\small b.) Structural rules for context manipulation}
\begin{equation*}
\tyrule{weak}
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{0} }} \vdash e : \sigma} 
\end{equation*}
\begin{equation*}
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
  {\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
\quad
\begin{array}{l}
\slift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
\slift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}
\begin{equation*}
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
~
\begin{array}{l}
\slift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
\slift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}

\caption{Structural coeffect liveness analysis}
\label{fig:applications-struct-bll}
\vspace{-1em}
\end{figure}

\paragraph{Bounded variable use.} 
The system discussed in this section tracks the number of times a variable is accessed in the 
call-by-name evaluation. Although we look at an example that tracks \emph{variable usage}, the same
system could be used to track access to resources that are always passed as a reference (and behave
effectively as call-by-name) and so the system is relevant for call-by-value languages too.
To demonstrate the idea, consider the following term:
%
\begin{equation*}
(\lambda v.x + v + v)~(x+y)
\end{equation*}
%
When evaluated, the body of the function directly accesses $x$ once and then twice indirectly, via
the function argument. Similarly, $y$ is accessed twice indirectly. Thus, the overall expression uses
$x$ three times and $y$ twice.

As discussed in Chapter~\ref{ch:structural}, the system preserves type and coeffect annotations under
the $\beta$-reduction. Reducing the expression in this case gives $x + (x+y) + (x+y)$. This has the 
same bounds as the original expression -- $x$ is used three times and $y$ twice.

\paragraph{Type system.}
The type system in Figure~\ref{fig:applications-struct-bll} annotates contexts with vectors of integers.
The rules have the same structure as those of the system for liveness analysis. The only difference is
how annotations are combined -- here, we use integer multiplication ($\ast$) and addition ($+$).

Variable access (\emph{var}) annotates a variable with $1$, meaning that it has been used once. An
unused variable (\emph{weak}) is annotated with $0$. Multiple occurrences of the same variable are 
introduced by contraction (\emph{contr}), which adds the numbers of the two contracted variables.

As previously, application (\emph{app}) and let binding (\emph{let}) combine two separate contexts.
The second part applies a function that uses its parameter $\cclrd{t}$-times to an argument that uses
variables in $\Gamma_2$ at most $\aclrd{\mathbf{s}}$-times (here, $\aclrd{\mathbf{s}}$ is a vector of
integers with an annotations for each variable in $\Gamma_2$). The sequential composition (modelling
call-by-name) multiplies the uses, meaning that the total number of uses is $(\cclrd{t} \ast \aclrd{\mathbf{s}})$
(where $\ast$ is a multiplication of a vector by a scalar). This models the fact that for each use of
the function parameter, we replicate the variable uses in $e_2$.

\paragraph{Example.} To type check the expression $(\lambda v.x+v+v)~(x+y)$ discussed earlier, we need
to use abstraction, application, but also the contraction rule. Assuming the type judgement for the body,
abstractions yields:
%
\begin{equation*}
\tyrule{abs}
 { \coctx{x\!:\!\mathbb{Z},v:\mathbb{Z}}{\alift{\cclrd{1},\cclrd{2}}} \vdash x+v+v : \mathbb{Z} }
 { \coctx{x\!:\!\mathbb{Z}}{\alift{\cclrd{1}}} \vdash (\lambda v.x+v+v) : \mathbb{Z} \xrightarrow{\cclrd{2}} \mathbb{Z} }
\end{equation*}
%
To type-check the application, the contexts of $e_1$ and $e_2$ need to contain disjoint variables.
For this reason, we $\alpha$-rename $x$ to $x'$ in the argument $(x+y)$ and later join $x$ and $x'$ using
the contraction rule. Assuming $(x'+y)$ is checked in a context that marks $x'$ and $y$ as used once, the 
application rule yields a judgement that is simplified as follows:
%
\begin{equation*}
\inference
  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}
          {\alift{\cclrd{1}} \cons (\cclrd{2} \ast \alift{\cclrd{1},\cclrd{1}}) } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
{\tyrule{contr}
  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{1}, \cclrd{2}, \cclrd{2}} } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
  { \coctx{x\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{3}, \cclrd{2}}} \vdash (\lambda v.x+v+v)~(x+y)  : \mathbb{Z}} }
\end{equation*}
%
The first step performs scalar multiplication, producing the vector
$\alift{\cclrd{1},\cclrd{2},\cclrd{2}}$. In the second step, we use contraction to join variables 
$x$ and $x'$ from the function and argument terms respectively.

\paragraph{Semantics.}
In the previous examples, we defined the semantics -- somewhat informally -- using a simple 
$\lambda$-calculus language to encode the model. More formally, this could be, for example,
a Cartesian closed category. In that model, we can reuse variables arbitrarily and so it is not
a good fit for modelling bounded reuse. Girard et al. \cite{logic-bounded} model their bounded
linear logic in an (ordinary) linear logic where variables can be used at most once.

Following the same approach, we could model a variable $\tau$, annotated with $\cclrd{r}$ as 
a product containing $\cclrd{r}$ copies of $\tau$, that is $\tau^{\cclrd{r}}$:
%
\begin{equation*}
\begin{array}{r}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
  ~:~ (\tau^{\cclrd{r_1}}_1 \times \ldots \times \tau^{\cclrd{r_n}}_n) \rightarrow \tau\\[0.5em]
\textnormal{where}~\tau^{\cclrd{r_i}}_i = \underbrace{\tau_i \times \ldots \times \tau_i}_{\cclrd{r_i}-\text{times}}
\end{array}
\end{equation*}
%
The functions are interpreted similarly. A function $\tau_1 \xrightarrow{\cclrd{t}} \tau_2$ is modelled
as a function taking $\cclrd{t}$-element product of $\tau_1$ values: $\tau_1^{\cclrd{t}} \rightarrow \tau_2$.

The rules that define the semantics of bounded calculus are mostly the same as (or easy to adapt 
from) the semantic rules of liveness in Figure~\ref{fig:applications-struct-livesem}. The ones that
differ are those that use sequential composition (application and let binding) and the contraction
rule, which represents pointwise composition. The bounded reuse calculus rules look as follows:

\begin{equation*}
\hspace{-0.5em}
\tag{\emph{app}}
\begin{array}{l}
\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\mathbf{v_2})
\end{array}
\end{equation*}

\begin{equation*}
\hspace{-0.5em}
\tag{\emph{contr}}
\begin{array}{l}
\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} =\\[-0.25em] 
\qquad  \lambda(\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{s}+\cclrd{t}}), \mathbf{v_2}) \rightarrow \\[-0.25em]
\qquad\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}\\[-0.25em]
\qquad\qquad\qquad (\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{t}}), (x_{\cclrd{t}+1}, \ldots, x_{\cclrd{t}+\cclrd{s}} ), \mathbf{v_2})
\end{array}
\end{equation*}

\paragraph{Bounded linear logic.}

Bounded linear logic \cite{logic-bounded} restricts well-typed terms to polynomial-time algorithms.  
This is done by limiting the number of times a value (proposition) can be used. An assumption
$!_k A$ means that a variable can be used at most $k$ times. 
We attach annotations to the whole context rather than individual assumptions and so a context
$!_{k_1} A_1, ..., !_{k_n} A_n$ is written as $\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$. 
This difference is further explained in Section~\ref{sec:related-work}. %For uniformity
%with later systems, we also write assumptions $A$ as $\tau$ and annotations $K$ as $\cclrd{r}$.

@@

\paragraph{Summary.}

We really need just: $\ast$, $+$, $1$, $0$, $\leq$

composing flat and structural - meta vs. language semantics
